---
title: "Bayes_rule"
author: "Damjan Stefanovski"
date: "June 28, 2017"
output: html_document
---

        
        #Bayesian analysis
        
# Bayesian statistics posits a prior on the parameter of interest
#All inferences are then performed on the distribution of the parameter given the data, called the posterior
#In general, $$ \mbox{Posterior} \propto \mbox{Likelihood} \times \mbox{Prior} $$
#Therefore (as we saw in diagnostic testing) the likelihood is the factor by which our prior beliefs 
 #are updated to produce conclusions in the light of the data

        
```{r}

library(manipulate)
pvals <- seq(0.01, 0.99, length = 1000)
manipulate(
        plot(pvals, dbeta(pvals, alpha, beta), type = "l", lwd = 3, frame = FALSE),
        alpha = slider(0.01, 10, initial = 1, step = .5),
        beta = slider(0.01, 10, initial = 1, step = .5)
)
```

## Credible intervals

A Bayesian credible interval is the Bayesian analog of a confidence interval
A $95%$ credible interval, $[a, b]$ would satisfy $$ P(p \in [a, b] | x) = .95 $$
The best credible intervals chop off the posterior with a horizontal line in the same way we did for likelihoods
These are called highest posterior density (HPD) intervals


```{r}
pvals <- seq(0.01, 0.99, length = 1000)
x <- 13; n <- 20
myPlot <- function(alpha, beta){
    plot(0 : 1, 0 : 1, type = "n", xlab = "p", ylab = "", frame = FALSE)
    lines(pvals, dbeta(pvals, alpha, beta) / max(dbeta(pvals, alpha, beta)), 
            lwd = 3, col = "darkred")
    lines(pvals, dbinom(x,n,pvals) / dbinom(x,n,x/n), lwd = 3, col = "darkblue")
    lines(pvals, dbeta(pvals, alpha+x, beta+(n-x)) / max(dbeta(pvals, alpha+x, beta+(n-x))),
        lwd = 3, col = "darkgreen")
    title("red=prior,green=posterior,blue=likelihood")
}
manipulate(
    myPlot(alpha, beta),
    alpha = slider(0.01, 10, initial = 1, step = .5),
    beta = slider(0.01, 10, initial = 1, step = .5)
    )


```

## Install the \texttt{binom} package, then the command



```{r}

library(binom)
binom.bayes(13, 20, type = "highest")
```


# The default credible level is $95%$ and the default prior is the Jeffrey's prior.

```{r}
pvals <- seq(0.01, 0.99, length = 1000)
x <- 13; n <- 20
myPlot2 <- function(alpha, beta, cl){
    plot(pvals, dbeta(pvals, alpha+x, beta+(n-x)), type = "l", lwd = 3,
    xlab = "p", ylab = "", frame = FALSE)
    out <- binom.bayes(x, n, type = "highest", 
        prior.shape1 = alpha, 
        prior.shape2 = beta, 
        conf.level = cl)
    p1 <- out$lower; p2 <- out$upper
    lines(c(p1, p1, p2, p2), c(0, dbeta(c(p1, p2), alpha+x, beta+(n-x)), 0), 
        type = "l", lwd = 3, col = "darkred")
}
manipulate(
    myPlot2(alpha, beta, cl),
    alpha = slider(0.01, 10, initial = 1, step = .5),
    beta = slider(0.01, 10, initial = 1, step = .5),
    cl = slider(0.01, 0.99, initial = 0.95, step = .01)
    )


```




